---
title: "Turing Patterns"
output: html_notebook
---


# Introduction

I've been fascinated by the patterns produced by [reaction-diffusion equations](https://en.wikipedia.org/wiki/Turing_pattern) (and the analogous chemical and biological systems) since I first heard about them many years ago - I've now implemented some of them in code. There are zillions of examples of this sort of thing online (although maybe not in *R*), but this is mine to play with!

```{r}
#Required libraries
library(tidyverse)
library(OpenImageR)
library(reshape2)
```

# Reaction Diffusion Equations (Gray-Scott)

These are beautifully descibed [here](http://www.karlsims.com/rd.html). I merely reiterate...

$$A'=A+(D_A\nabla^2A-AB^2+f(1-A))\Delta t$$
$$B'=B+(D_B\nabla^2B+AB^2-(k+f)B)\Delta t$$

The terms are:

- $A$ the amount of reactant A
- $B$ the amount of reactant B
- $D_A$, $D_B$ the diffusion coefficients of the reactants
- $f$ the feed - this describes the amount of raectant being added to the system
- $k$ the kill - this descibes the amount of reactant B being removed from the system

Breaking the equations down, as I understand, the terms containing the diffusion coefficients are basically just [Fick's second law](https://en.wikipedia.org/wiki/Fick%27s_laws_of_diffusion) describing how things diffuse. The $AB^2$ terms are about the reaction between A and B and the final terms are to do with reactant being added or removed from the system.

# R implementation

Both of these equations are implemented by R functions. The Laplacian ($\nabla^2$) operator is approximated by a convolution.

```{r}
laplacian_kernel=matrix(c(.05,.2,.05,.2,-1,.2,.05,.2,.05),nrow=3,ncol=3)
a_prime<-function(A,B,D_A,f,l_kernel){
  L_A=convolution(A,l_kernel)
  A+(D_A*L_A-A*B*B+f*(1-A))
}
b_prime<-function(A,B,D_B,f,k,l_kernel){
  L_B=convolution(B,l_kernel)
  B+(D_B*L_B+A*B*B-(k+f)*B)
}
```

The initial conditions and constants are then established:

```{r}

set.seed(1) #Makes "random" number generation repeatable

size=500 #Image size

#Set up a matrix for reactant A - each cell contains the concentration of the reactant in that region of space
A=matrix(1,nrow=size,ncol=size)
rownames(A)=1:size
colnames(A)=1:size

#Set up matrix for reactant B
B=matrix(0,nrow=size,ncol=size)
B[round(0.45*size):round(0.55*size),round(0.45*size):round(0.55*size)]=1
rownames(B)=1:size
colnames(B)=1:size

#Initialise constants
D_A=1
D_B=.5
f=.055
k=.062

```

The functions are then iterated (ideally, generally, we avoid loops in R, but this one is hard to get round). I'm using a data frame to store every 2000th iteration so I can show how the pattern develops over time...

```{r}
AB_series<-data.frame()
block_size<-2000
start_time<-Sys.time()
for(i in 0:5){
  #Take "snapshots" of patterns evey block_size iterations
  AB_series<-AB_series %>% bind_rows(
    bind_rows(
      melt(A) %>% add_column(component="A"),
      melt(B) %>% add_column(component="B")
    ) %>% add_column(iteration=i*block_size))
  #Iterate the equations
  for(j in 1:block_size){
    A<-a_prime(A,B,D_A,f,laplacian_kernel)
    B<-b_prime(A,B,D_B,f,k,laplacian_kernel)
  }
}
end_time<-Sys.time()
end_time-start_time
```

As you can see, this takes ages, but if I'm doing loops in *R* I suppose I deserve everything I get!

The data frame containing every 2000th iteration is then plotted with *ggplot*:

```{r, fig.height=3, fig.width=9}

AB_series %>%
  ggplot(aes(x = Var2, y = Var1)) + 
    geom_raster(aes(fill=value))+
    coord_fixed()+
    facet_grid(paste("Reactant",component)~iteration)+
    scale_fill_viridis_c()+
    labs(x="x",y="y",fill="Amount", subtitle = "Iteration")+
    theme(text = element_text(size=16))
  


```

I love how such complex and unexpected patterns emerge from a fairly simple setup. I may come back to this and have a play with the parameters.

See my [website](https://sites.google.com/view/robert-turner) for more code and stuff. Contact me on bob dot turner dot uk at gmail dot com with comments or questions.

